---
title: "PCA of SNP data"
author: "Alayna Mead"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
    highlight: kate
---

Plot PCAs from processed SNP data using vegan

# Setup

## packages and data

```{r setup}

library('vegan') # redundancy analysis
library('adegenet') # read.PLINK()
library('psych') # pairs.panels()

# Following:
#https://popgen.nescent.org/2018-03-27_RDA_GEA.html

# site data - named dat, needs to be renamed
load('data/clean/CCGP_samples_channel_islands_clean.rda')
info <- dat

# use this to read SNPs!
dat <- read.PLINK('data/raw/Qtom107.Qchr17.Qssp3.20220906.qlob.ef.repeatsOut.renamedChrsVars.biallelicSNPs.meanDP5.genoDP5.MAF0.01.missing0.9.ldPruned.additive.raw', 
                  n.cores = 2, 
                  quiet = F)

sessionInfo()


```

```{r markdown_settings}

knitr::opts_chunk$set(fig.width = 10, fig.height = 8)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)

```

## setup SNP data

```{r snp_data_setup}

# look at dataset
# it omits the non-SNP info
dat
dim(dat)

snps <-  as.matrix(dat)
snps[1:10, 1:10]
dim(snps)

# optional - get a subset of SNPs for faster analysis, then run on whole dataset
#snps <- snps[, sample(1:ncol(snps), 10000, replace = F)]

```


## setup collection data

```{r data_setup}

# get info only for the ones we have SNP data for

# first check for any mismatches
rownames(snps)[! rownames(snps) %in% info$ID_vcf]

# get subset
info <- info[match(rownames(snps), info$ID_vcf),]

# check
cbind(rownames(snps), info$ID_vcf, rownames(info))

# rename SNP rows to match corrected names (info rownames)
rownames(snps) <- rownames(info)

# also get species abbreviation from name
info$sp <- sapply(1:nrow(snps), function(x) strsplit(rownames(snps)[x], split = '.', fixed = T)[[1]][1])
info$sp <- factor(info$sp)



```

# impute SNP data

```{r impute}

# need to impute missing data

# proportion missing data - about 5%
sum(is.na(snps))/length(unlist(snps))

# missing data by individual
sapply(1:nrow(snps), function(x) sum(is.na(snps[x,])))
miss <- sapply(1:nrow(snps), function(x) sum(is.na(snps[x,]))/ncol(snps))
missdf <- cbind(rownames(snps), miss)
missdf[order(missdf[,2]),]


# impute by most common snp
imp <- apply(snps, 2, function(x) replace(x, is.na(x), as.numeric(names(which.max(table(x))))))
sum(is.na(imp))


```

```{r check}

# one last check that rows are in the same order...
cbind(rownames(info), rownames(snps), rownames(imp))

```


# Plots

## Simple PCA, all samples

Uses vegan rda() function without constraining matrices, which just runs a PCA

From documentation: "If both matrices Z and Y are missing, the data matrix is analysed by ordinary correspondence analysis (or principal components analysis)."

```{r PCA_all_samples}

# run the redundancy analysis

rda <- rda(imp, scale = T)

rda
#summary(rda)
screeplot(rda)

# save summary
rda.info <- summary(rda)
rda.info$sites

# plot the individuals ('sites')

# set params
par(cex.axis = 1.2, cex.lab = 1.5)
par(mar = c(5,5,3,1))

# setup colors and points

# color by island
bg <- c('black', "#3c4a8b","#009c85","#84bc5f","#edb829","#f57404","#b30000")

# point shape by species
sp <- c(23, 22, 21)


#png(filename = 'results/pca/PCA_allSamples_axes1-2.png', res = 300, height = 5.5, width = 6, units = 'in')
#pdf(file = 'results/pca/PCA_allSamples_axes1-2.pdf', height = 5.5, width = 6)

par(cex.axis = 1.2, cex.lab = 1.5)
par(mar = c(5,5,3,1))

# PC1 and PC2
choices = c(1,2)

# x and ylab code is ugly, but it's just pulling the variance explained from the RDA summary for each axis
plot(rda, 
     type = 'n', 
     choices = choices,
     xlab = paste('PC', choices[1], ' (', round(rda.info$cont$importance[2, choices[1]], 3)*100, '% variance explained)', sep = ''),
     ylab = paste('PC', choices[2], ' (', round(rda.info$cont$importance[2, choices[2]], 3)*100, '% variance explained)', sep = ''))

points(rda, pch = sp[info$sp], cex = 1.5, display = 'sites', bg = bg[info$island], col = 'black', choices = choices)

legend('topleft', legend = c(levels(info$sp), levels(info$island)), pch = c(sp, rep(15, 8)), col = c(rep('black', 3), bg), cex = 0.8)

#dev.off()

# PC1 and PC3

#png(filename = 'results/pca/PCA_allSamples_axes1-3.png', res = 300, height = 5.5, width = 6, units = 'in')
#pdf(file = 'results/pca/PCA_allSamples_axes1-3.pdf', height = 5.5, width = 6)

par(cex.axis = 1.2, cex.lab = 1.5)
par(mar = c(5,5,3,1))

choices = c(1,3)
plot(rda, 
     type = 'n', 
     choices = choices,
     xlab = paste('PC', choices[1], ' (', round(rda.info$cont$importance[2, choices[1]], 3)*100, '% variance explained)', sep = ''),
     ylab = paste('PC', choices[2], ' (', round(rda.info$cont$importance[2, choices[2]], 3)*100, '% variance explained)', sep = ''))

points(rda, pch = sp[info$sp], cex = 1.5, display = 'sites', bg = bg[info$island], col = 'black', choices = choices)

#legend('topleft', legend = c(levels(info$sp), levels(info$island)), pch = c(sp, rep(15, 8)), col = c(rep('black', 3), bg), cex = 1.2)

#dev.off()


```


## Only Channel Islands, exclude Guadalupe Island

```{r rda_subset_california}

# plot without guadalupe island

imp.sub <- imp[! info$island %in% c("Guadalupe Island"),]
info.sub <- info[! info$island %in% c("Guadalupe Island"),]

rda <- rda(imp.sub, scale = T)

rda
#summary(rda)
screeplot(rda)

# save summary
rda.info <- summary(rda)
rda.info$sites

# plot the individuals ('sites')

# set params
par(cex.axis = 1.2, cex.lab = 1.5)
par(mar = c(5,5,3,1))

# setup colors and points

# color by island
bg <- c('black', "#3c4a8b","#009c85","#84bc5f","#edb829","#f57404")

# point shape by species
sp <- c(23, 22, 21)


# png(filename = 'results/pca/PCA_California_axes1-2.png', res = 300, height = 5.5, width = 6, units = 'in')
# pdf(file = 'results/pca/PCA_California_axes1-2.pdf', height = 5.5, width = 6)


par(cex.axis = 1.2, cex.lab = 1.5)
par(mar = c(5,5,3,1))

# PC1 and PC2
choices = c(1,2)

# x and ylab code is ugly, but it's just pulling the variance explained from the RDA summary for each axis
plot(rda, 
     type = 'n', 
     choices = choices,
     xlab = paste('PC', choices[1], ' (', round(rda.info$cont$importance[2, choices[1]], 3)*100, '% variance explained)', sep = ''),
     ylab = paste('PC', choices[2], ' (', round(rda.info$cont$importance[2, choices[2]], 3)*100, '% variance explained)', sep = ''))

points(rda, pch = sp[info.sub$sp], cex = 1.5, display = 'sites', bg = bg[info.sub$island], col = 'black', choices = choices)

#legend('topleft', legend = c(levels(info$sp), levels(info$island)), pch = c(sp, rep(15, 8)), col = c(rep('black', 3), bg), cex = 0.8)

#dev.off()

# PC1 and PC3

# png(filename = 'results/pca/PCA_California_axes1-3.png', res = 300, height = 5.5, width = 6, units = 'in')
# pdf(file = 'results/pca/PCA_California_axes1-3.pdf', height = 5.5, width = 6)

par(cex.axis = 1.2, cex.lab = 1.5)
par(mar = c(5,5,3,1))

choices = c(1,3)
plot(rda, 
     type = 'n', 
     choices = choices,
     xlab = paste('PC', choices[1], ' (', round(rda.info$cont$importance[2, choices[1]], 3)*100, '% variance explained)', sep = ''),
     ylab = paste('PC', choices[2], ' (', round(rda.info$cont$importance[2, choices[2]], 3)*100, '% variance explained)', sep = ''))

points(rda, pch = sp[info.sub$sp], cex = 1.5, display = 'sites', bg = bg[info.sub$island], col = 'black', choices = choices)

#legend('topleft', legend = c(levels(info$sp), levels(info$island)), pch = c(sp, rep(15, 8)), col = c(rep('black', 3), bg), cex = 1.2)

#dev.off()

```

## Only Northern Channel Islands

```{r rda_subset_northern_islands}

# plot just SRI and SCI for easier comparison with Qagr data

imp.sub <- imp[info$island %in% c("Mainland", "Santa Rosa Island", "Santa Cruz Island"),]
info.sub <- info[info$island %in% c("Mainland", "Santa Rosa Island", "Santa Cruz Island"),]

rda <- rda(imp.sub, scale = T)

rda
#summary(rda)
screeplot(rda)

# plots

# setup colors and points

# color by island
bg <- c('black', "#3c4a8b","#009c85","#84bc5f","#edb829","#f57404","#b30000")

# point shape by species
sp <- c(23, 22, 21)
#sp <- c(5, 0, 1)

# PC1 and PC2
#png(file = 'results/redundancy_analysis/rda_plot_indivs_clim_Qtom107.Qchr17.Qssp3_SantaRosa_SantaCruz_Mainland_PC1_PC2.png', height = 6, width = 6, res = 300, units = 'in')
choices = c(1,2)
plot(rda, type = 'n', choices = choices)
points(rda, pch = sp[info.sub$sp], cex = 1, display = 'sites', bg = bg[info.sub$island], col = 'black', choices = choices)
#points(rda, pch = sp[info.sub$sp], cex = 2, display = 'sites', col = bg[info.sub$island], choices = choices)
legend('bottomleft', legend = c(levels(info.sub$sp), "Mainland Qchr", "Santa Rosa Island", "Santa Cruz Island"), pch = c(sp, rep(15, 8)), col = c(rep('black', 4), "#3c4a8b","#009c85"), cex = 1)
#dev.off()


# PC1 and PC3
#png(file = 'results/redundancy_analysis/rda_plot_indivs_clim_Qtom107.Qchr17.Qssp3_SantaRosa_SantaCruz_Mainland_PC1_PC3.png', height = 6, width = 6, res = 300, units = 'in')
choices = c(1,3)
plot(rda, type = 'n', choices = choices)
points(rda, pch = sp[info.sub$sp], cex = 1, display = 'sites', bg = bg[info.sub$island], col = 'black', choices = choices)
#dev.off()

###########################
# no mainland

imp.sub <- imp[info$island %in% c("Santa Rosa Island", "Santa Cruz Island"),]
info.sub <- info[info$island %in% c("Santa Rosa Island", "Santa Cruz Island"),]

rda <- rda(imp.sub, scale = T)

rda
#summary(rda)
screeplot(rda)

# plots

# setup colors and points

# color by island
bg <- c('black', "#3c4a8b","#009c85","#84bc5f","#edb829","#f57404","#b30000")

# point shape by species
sp <- c(23, 22, 21)
#sp <- c(5, 0, 1)

# PC1 and PC2
#png(file = 'results/redundancy_analysis/rda_plot_indivs_clim_Qtom107.Qchr17.Qssp3_SantaRosa_SantaCruz_PC1_PC2.png', height = 6, width = 6, res = 300, units = 'in')
choices = c(1,2)
plot(rda, type = 'n', choices = choices)
points(rda, pch = sp[info.sub$sp], cex = 1, display = 'sites', bg = bg[info.sub$island], col = 'black', choices = choices)
#points(rda, pch = sp[info.sub$sp], cex = 2, display = 'sites', col = bg[info.sub$island], choices = choices)
legend('topleft', legend = c(levels(info.sub$sp), "Mainland Qchr", "Santa Rosa Island", "Santa Cruz Island"), pch = c(sp, rep(15, 8)), col = c(rep('black', 3), "#3c4a8b","#009c85"), cex = 1)
#dev.off()


# PC1 and PC3
#png(file = 'results/redundancy_analysis/rda_plot_indivs_clim_Qtom107.Qchr17.Qssp3_SantaRosa_SantaCruz_PC1_PC3.png', height = 6, width = 6, res = 300, units = 'in')
choices = c(1,3)
plot(rda, type = 'n', choices = choices)
points(rda, pch = sp[info.sub$sp], cex = 1, display = 'sites', bg = bg[info.sub$island], col = 'black', choices = choices)
#dev.off()

# combined fig

#png(file = 'results/redundancy_analysis/rda_plot_indivs_clim_Qtom107.Qchr17.Qssp3_SantaRosa_SantaCruz_PC1-PC3.png', height = 6, width = 12, res = 300, units = 'in')
par(mfrow = c(1,2))
choices = c(1,2)
plot(rda, type = 'n', choices = choices, main = substitute(paste(italic("Q. tomentella"))))
points(rda, pch = sp[info.sub$sp], cex = 1, display = 'sites', bg = bg[info.sub$island], col = 'black', choices = choices)

legend('bottomright', legend = c(levels(info.sub$sp), "Santa Rosa Island", "Santa Cruz Island"), pch = c(sp, rep(15, 8)), col = c(rep('black', 3), "#3c4a8b","#009c85"), cex = 0.8)


# PC1 and PC3
choices = c(1,3)
plot(rda, type = 'n', choices = choices)
points(rda, pch = sp[info.sub$sp], cex = 1, display = 'sites', bg = bg[info.sub$island], col = 'black', choices = choices)

#dev.off()

```